Computing with an Angry Poker Player

### 2014-09-25 draft0.tex
Edited Section 2
Fixed grammar and notation errors in Section 2.

\vbox{
\noindent
\begin{quote}
Each square of an infinite chess board contains some nonnegative number of poker chips. An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions. 
\end{quote}
}
We also consider extensions of this problem set up where the earthquake hits all squares on the board at once. We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
\section{Standard Board}

Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz

@@

\begin{definition}
 We define the small earthquake operator as a map from the set of all boards and locations to the set of all boards and write $SE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
If $f(x,y) \leq 4$, then $SE(f, (x,y)) = f$, otherwise,  
\begin{align*}
SE( f(x,y), (x,y)) = f(x,y)-4 \\
SE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
SE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
SE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
SE( f(x,y-1), (x,y)) = f(x,y-1)+1,
\end{align*}
and leaves all other squares unchanged.
\end{definition}

@@

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
\end{theorem}

We will first show that finite boards cannot return to a previous state. We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy. This shows that all finite boards

So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $SE^{M}(f) = SE^{M+k}(f)$

@@

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE(f)) - \Phi(f)$.
\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (|x|+|y|)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.

\begin{proof}
We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
\begin{align}
\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
\end{align}
And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
\begin{align}
4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
\end{align}
\end{proof}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$. Then by Lemma~\ref{finiteextension}, we have that all $E^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.

There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $E^N(f) = E^{N+m}(f)$ for all $m \geq 0$.
\end{proof}

\section{$n$-Trees}
Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.

>>>>

\vbox{
\noindent
\begin{quote}
Each square of an infinite chess board contains some nonnegative number of poker chips. An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions. 
\end{quote}
}
We also consider extensions of this problem set up where the earthquake hits all squares on the board at once. We refer to the former type of earthquake as $\textbf{small earthquakes}$ and earthquakes that hit the entire board as $\textbf{big earthquakes}$.
\section{Standard Board}

Written by Erik Waingarten and Perry Kleinhenz, edited by Fermi Ma

@@

\begin{definition}
 We define the small earthquake operator as a map from the set of all boards and locations to the set of all boards and write $SE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
If $f(x,y) \leq 4$, then $SE(f, (x,y)) = f$, otherwise,  
\begin{align*}
SE( f(x,y), (x,y)) = f(x,y)-4 \\
SE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
SE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
SE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
SE( f(x,y-1), (x,y)) = f(x,y-1)+1,
\end{align*}
and leaves all other squares unchanged.

We use the notation $SE^n(f, \{(x_i,y_i)\})$ to refer to the application of $SE$ to the board $f$, $n$ times to the squares $(x_1, y_1), \ldots (x_{n}, y_{n})$ respectively. 
\end{definition}

@@

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ such that after $N$ big earthquakes or $N$ small earthquakes the resultant board will be stable.
\end{theorem}

We will first show that finite boards cannot return to a previous state. We will then show that for a given finite board there is a maximum nunmber of states that a board can occupy. This shows that all finite boards eventually become stable.

A restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and $SE^{M}(f) = SE^{M+k}(f)$

@@

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
That is if big earthquake changes the state of the board, then $\Phi(f)$ increases, therefore if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or in other words that you never return to your original state.

We can evaluate $\Phi(BE(f)) - \Phi(f)$.
\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (|x|+|y|)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]

Suppose $f(x_0,y_0) \geq 4$ for some location $(x_0, y_0)$, then the number of chips in location $(x_0,y_0)$ will have decreased by $4$ after applying $BE$. 
For each decrease in the number of chips at location $(x_0,y_0)$ of $4$, the number of chips at the location $(y,z)$ increases by $1$, for each $(y,z) \in n(x_0,y_0)$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(BE(f)) - \Phi(f) = \sum_{(x,y) \in A} \left((-4)(|x|+|y|)^2 + \sum_{(v,w) \in n(x,y)} (|v| + |w|)^2 \right) \]
\[\geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]

\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
where the first inequality holds because at most two of the neighbors of $(x,y)$ are one unit closer to the origin then $(x,y)$ and at least two are one unit farther away and $(|x|+|y|+1)^2 > 2(|x|+|y|-1)^2$
\end{proof}

\begin{lemma}
If $SE(f, (x_1, y_1)) \neq f$, then $SE^n(f, \{(x_n,y_n)\}) \neq f$ for all $n > 0$ and all sequences $\{(x_n, y_n)\}$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.

\begin{proof}
We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
and evaluate $\Phi(SE(f, (x_1,y_1))) - \Phi(f)$:
\begin{align}
\Phi(SE(f), (x_1, y_1)) - \Phi(f) =\sum_{x,y} SE(f(x,y), (x_1, y_1))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
= SE(f(x_1,y_1), (x_1, y_1))(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1,y_1)} SE(f(x,y), (x_1, y_1))(|x|+|y|)^2 \\- f(x_1,y_1)(|x_1|+|y_1|)^2 + \sum_{(x,y) \in n(x_1, y_1)} f(x,y)(|x|+|y|)^2 \\
= 4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2
\end{align}
And because at least two of the neighbors of $(x_1, y_1)$ are one unit farther from the origin than $(x_1, y_1)$ and at most two of the neighbors are one unit closer, and $(|x_1|+|y_1| +1)^2 > (|x_1|+|y_1| - 1)^2$, we can see,
\begin{align}
4(|x_1| + |y_1|)^2 +\sum_{(x,y) \in n(x_1, y_1)} (|x| + |y|)^2  \\ 
\geq 4(|x_1| + |y_1|)^2 + 2 (|x_1| + |y_1|+1)^2 + 2 (|x_1| + |y_1|-1)^2 = 4 >0
\end{align}
\end{proof}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}_f$, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie within the locations $[x_l, x_h] \times [y_l, y_h]$. Then by Lemma~\ref{finiteextension}, we have that for all $n$ all chips on $SE^n(f)$ and $BE^n(f)$ will lie in $R = [x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l - \sqrt{S}, y_h + \sqrt{S}]$.

There exists less than $(x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ possible boards $b \in \mathcal{B}_f$ that satisfy this. We know that there are no cycles in the states, so there must be an $0 \leq N < (x_h - x_l + 2\sqrt{S})(y_h - y_l + 2\sqrt{S})S$ such that $SE^N(f) = SE^{N+m}(f)$ and $BE^N(f)=BE^{N+m}$ for all $m \geq 0$.
\end{proof}

\section{$n$-Trees}
Written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten

Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.
